Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-27 paper-new/concrete-model.tex
more squeezing
12 pages

Now we define the perturbations for $D$.
Recall from our construction of a model with perturbations
(Section \ref{sec:constructing-perturbations}) that for each value
type $A$ we associate a monoid $P_A$ of perturbations
and a homomorphism into the monoid of endomorphisms bisimilar to the identity,
and likewise for computation types.
We define the perturbations for $D$ via least-fixpoint in the category of monoids as
%
\[ P_D \cong (P_{D \times D}) \times P_{U(D \to FD)}. \]
%
Unfolding the above, the above equation is the same as
%
\[ P_D \cong (P_D \times P_D) \times (\mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D). \]
%
We now explain how to interpret these perturbations as endomorphisms.
We define $\ptb_D : P_D \to \{ f : D \to D \mid f \bisim \id \}$ below,
% via the universal property of the coproduct of monoids, giving a case for each of the generators.
% In the below, note the use of the functorial action of $\arr$ on morphisms.

%
% \[ P_D \cong (P_D \times P_D) \times (P_D \times (\mathbb{N} \times P_D)), \]
%

$\ptb_D((p_{\text{times}}), p_{\text{fun}}) = \lambda d.\text{case $d$ of}$
\begin{align*}
    &\alt \tnat(m) \mapsto \tnat(m) \\
    &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(\ptb_{D \times D}(p_\text{times})(d_1, d_2))} \\
    &\alt \tfun(\tilde{f}) \mapsto {\tfun(\lambda t. \ptb_{U(D \to FD)}()(\tilde{f}_t))} \\
\end{align*}

    % \item $\ptb_D(1)$
    % \item $\ptb_D(\delta^K_D)$ is defined similarly to the previous but has
    % \[ \id \arr i^K(\delta^K_D) \] instead.
    % \item $\ptb_D(\delta^K_D) = \lambda d.\text{case $d$ of}$
    %   \begin{align*} 
    %     &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(i^K(\delta^K_D)(d_1), d_2)} \\
    %     &\alt d' \to d'
    %   \end{align*}
    % \item $\ptb_D(\delta^K_D)$ is defined similarly to the previous but has 
    % \[ (d_1, \ptb_D(\delta^K_D)(d_2)) \] instead.

One can verify that this forms a homomorphism from $P_D \to \{ f : D \to D : f \bisim \id \}$.
We claim that the three relations $\inat$, $\itimes$, and $\iarr$
%and their lifted versions 
satisfy the push-pull property.
As an illustrative case, we establish the push-pull property for the relation $\iarr$.
We define $\pull_{\iarr} : P_D \to P_{U(D \arr FD)}$ by
%
\[ \pull_{\iarr}(p_{\text{times}}, p_{\text{fun}}) = p_{\text{fun}}, \]
% (recall that $P_{U(D \arr FD)} = \mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D$).
%
i.e., we simply forget the other perturbation.
%
We define $\push_{\iarr} : P_{U(D \arr FD)} \to P_D$ by
%
\[ \push{\iarr}(p_{\text{fun}}) = (\id, p_{\text{fun}}), \]
%
Showing that the relevant squares commute is straightforward.
%
% To do so, consider an arbitrary perturbation $p$ on $D$. 
% Let $(f, d) \in \iarr$. This means that $d$ must be of the form $\tfun{\tilde{f}}$. 
% When we interpret the perturbation on $D$, obtaining and endomorphism to which we then apply $d$,
% we will be perform the action that was performed by the perturbation on the other side and thus
% we will be done by our assumption that $f$ is related to $d$.

We next claim that the relations $\inat$, $\itimes$, and $\iarr$ are quasi-left-representable,
and that their lifts are quasi-right-representable.
Indeed, since the relations are functional, it is easy to see that they are quasi-left-representable
where the perturbations are taken to be the identity.

For quasi-right-representability, the most interesting case is $\li(\iarr)$.
Defining the projection $p_{\iarr} : FD \to FU(D \to FD)$ is equivalent to defining
$p' : D \to UFU(D \to FD)$. We define

$p' = \lambda d.\text{case $d$ of}$
\begin{align*}
    &\alt \tnat(m) \mapsto \mho \\
    &\alt \ttimes(d_1, d_2) \mapsto \mho \\
    &\alt \tfun(\tilde{f}) \mapsto \theta (\lambda t. \eta(\tilde{f}_t)).
\end{align*}

We define $\dellp_D = \delrp_D = \delta_{FD}$.
Then it is easy to show using the definition of $\iarr$ that the squares for $\dnl$ and $\dnr$ commute.
% Then for $\dnr$, we need to show that if $(f, d) \in \iarr$ then

It is also straightforward to establish the retraction property for
each of these three relations. In the case of $\iarr$, we
have that the property holds up to a delay.

@@

Given a predomain $X$ on a clock-irrelevant type, we can define a
``global'' version of the lock-step error ordering and the
weak bisimilarity relation on elements of the global lift; the former is defined by
%
\[ x \le^{gl}_X y := \forall k. x[k] \le y[k], \]
%
and the latter is defined by
%
\[ x \bisim^{gl}_X y := \forall k. x[k] \bisim y[k]. \]
%
On the other hand, we can define coinductively a ``lock-step error ordering"
relation on $\delay(X + 1)$:
%
\begin{mathpar}
  \inferrule*[]
  { }
  {\tnow (\inr\, 1) \ledelay d}

>>>>

Now we define the perturbations for $D$.
Recall from our construction of a model with perturbations
(Section \ref{sec:constructing-perturbations}) that for each value
type $A$ we associate a monoid $P_A$ of perturbations
and a homomorphism into the monoid of endomorphisms bisimilar to the identity,
and likewise for computation types.
We define the perturbations for $D$ via least-fixpoint in the category of monoids as
%
\( P_D \cong (P_{D \times D}) \times P_{U(D \to FD)}. \)
%
Unfolding the definitions, this is the same as
%
\( P_D \cong (P_D \times P_D) \times (\mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D). \)
%
We now explain how to interpret these perturbations as endomorphisms.
We define $\ptb_D : P_D \to \{ f : D \to D \mid f \bisim \id \}$ below,
% via the universal property of the coproduct of monoids, giving a case for each of the generators.
% In the below, note the use of the functorial action of $\arr$ on morphisms.
%
% \[ P_D \cong (P_D \times P_D) \times (P_D \times (\mathbb{N} \times P_D)), \]
%
\begin{align*}
 \ptb_D(p_{\text{times}}, p_{\text{fun}}) &= \lambda d.\text{case $d$ of}  \\
 &\alt \tnat(m) \mapsto \tnat(m) \\
    &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(\ptb_{D \times D}(p_\text{times})(d_1, d_2))} \\
    &\alt \tfun(\tilde{f}) \mapsto {\tfun(\lambda t. \ptb_{U(D \to FD)}()(\tilde{f}_t))} \\
\end{align*}
    % \item $\ptb_D(1)$
    % \item $\ptb_D(\delta^K_D)$ is defined similarly to the previous but has
    % \[ \id \arr i^K(\delta^K_D) \] instead.
    % \item $\ptb_D(\delta^K_D) = \lambda d.\text{case $d$ of}$
    %   \begin{align*} 
    %     &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(i^K(\delta^K_D)(d_1), d_2)} \\
    %     &\alt d' \to d'
    %   \end{align*}
    % \item $\ptb_D(\delta^K_D)$ is defined similarly to the previous but has 
    % \[ (d_1, \ptb_D(\delta^K_D)(d_2)) \] instead.
One can verify that this defines a homomorphism from $P_D \to \{ f : D \to D : f \bisim \id \}$.
We claim that the three relations $\inat$, $\itimes$, and $\iarr$
%and their lifted versions 
satisfy the push-pull property.
As an illustrative case, we establish the push-pull property for the relation $\iarr$.
We define $\pull_{\iarr} : P_D \to P_{U(D \arr FD)}$ by\( \pull_{\iarr}(p_{\text{times}}, p_{\text{fun}}) = p_{\text{fun}}, \)
% (recall that $P_{U(D \arr FD)} = \mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D$).
%
i.e., we simply forget the other perturbation.
%
We define $\push_{\iarr} : P_{U(D \arr FD)} \to P_D$ by \( \push{\iarr}(p_{\text{fun}}) = (\id, p_{\text{fun}}) \).
%
Showing that the relevant squares commute is straightforward.
%
% To do so, consider an arbitrary perturbation $p$ on $D$. 
% Let $(f, d) \in \iarr$. This means that $d$ must be of the form $\tfun{\tilde{f}}$. 
% When we interpret the perturbation on $D$, obtaining and endomorphism to which we then apply $d$,
% we will be perform the action that was performed by the perturbation on the other side and thus
% we will be done by our assumption that $f$ is related to $d$.

We next claim that the relations $\inat$, $\itimes$, and $\iarr$ are quasi-left-representable,
and that their lifts are quasi-right-representable.
Indeed, since the relations are functional, it is easy to see that they are quasi-left-representable
where the perturbations are taken to be the identity.
%
For quasi-right-representability, the most interesting case is $\li(\iarr)$.
Defining the projection $p_{\iarr} : FD \to FU(D \to FD)$ is equivalent to defining
$p' : D \to UFU(D \to FD)$. We define
\begin{align*}
 p' = \lambda d.\text{case $d$ of}   &\alt \tnat(m) \mapsto \mho \\
    &\alt \ttimes(d_1, d_2) \mapsto \mho \\
    &\alt \tfun(\tilde{f}) \mapsto \theta (\lambda t. \eta(\tilde{f}_t)).
\end{align*}
%
We define $\dellp_D = \delrp_D = \delta_{FD}$.
Then it is easy to show using the definition of $\iarr$ that the squares for $\dnl$ and $\dnr$ commute.
% Then for $\dnr$, we need to show that if $(f, d) \in \iarr$ then
%
It is also straightforward to establish the retraction property for
each of these three relations. In the case of $\iarr$, we
have that the property holds up to a delay.

@@

Given a predomain $X$ on a clock-irrelevant type, we can define a
``global'' version of the lock-step error ordering and the
weak bisimilarity relation on elements of the global lift; the former is defined by
%
\( x \le^{gl}_X y := \forall k. x[k] \le y[k], \)
%
and the latter is defined by
%
\( x \bisim^{gl}_X y := \forall k. x[k] \bisim y[k]. \)
%
On the other hand, we can define coinductively a ``lock-step error ordering"
relation on $\delay(X + 1)$:
%
\begin{mathpar}
  \inferrule*[]
  { }
  {\tnow (\inr\, 1) \ledelay d}
